---
title: "SH & AU Tests"
author: "Ian G. Brennan"
output: pdf_document
---

# Shimodaira-Hasegawa and Approximately Unbiased Analyses
## Prerequisites:
1. Alignment (Phylip format for RAxML)
2. Partition File (optional)
3. Constraint Trees (monophyletic groups for topology testing). Make these yourself, and format them similar to those presented below:
```{r eval=FALSE}
((X,Y),A,B,C,D)
#this constrains X&Y, but leaves ABCD free)
((A,B),X,Y,C,D)
#this constrains A&B, but leaves XYCD free)
(((A,B),(X,Y)),C,D)
#this constrains A&B, X&Y, but also AB&XY, CD are always free
```
4. Your preferred ML topology - likely your unconstrained RAxML tree

## Necessary Programs:
1. RAxML: suggest downloading raxmlHPC-PTHREADS
    +compile our download from http://sco.h-its.org/exelixis/web/software/raxml/hands_on.html
2. CONSEL: get it from http://www.sigmath.es.osaka-u.ac.jp/shimo-lab/prog/consel/
    +use instructions from: http://phylo.bio.ku.edu/slides/lab7-Testing/lab7Testing.html

+before each step you'll have to physically drop the raxmlHPC-PTHREADS-SSE3 into the terminal  
+remember if you publish constrained trees, indicated constrained nodes with a unique symbol  
+things within square brackets [...] represent objects you're dragging/dropping  

\pagebreak

# Step by Step:
## Getting the Best Unconstrained Tree
Basic Glossary, other can be found in the RAxML Manual:  
-m [your molecular model of choice, I will use GTRGAMMA]  
-q [your partition file, this is optional]  
-s [your phylip alignment]  
-n [your output file name]  
+don't actually use the brackets

1. Run RAxML for 100 (or however many you want) searches for the best scoring unconstrained tree (command -N)
```{bash eval=FALSE}
[drag/drop RAxML executable] -T 8 -f d -N 100 -m [...] -q [...] -s [...] -n [...]
#above is an out line it will actually look something like this:
raxmlHPC-PTHREADS-SSE3 -T 8 -f d -N 100 -m GTRGAMMA -q partfile.txt -s align.phy -n Tree.tre
```
2. Run RAxML for 1000 (or however many you want) bootstraps (command -# and -b)
```{bash eval=FALSE}
[RAxML] -T 8 -f d -# 1000 -m [...] -q [...] -s [...] -n [...] -b 12345
# you can change -b to -x for rapid bootstrapping
```
3. Add your 1000 bootstraps to your best unconstrained tree (command -t, -z)
```{bash eval=FALSE}
[RAxML] -T 8 -f b -m [...] -q [...] -s [...] -z [...] -t [best tree, step1] -n [...]
```
4. Step 3 should culminate in producing a single best tree with bootstrap support values, check it in figtree to make sure you constrained the correct nodes.

## Getting the Best Tree for Each Constrained Scenario
1. We need to repeat the above steps (with one big change, constraints). We can constrain either using the -g or the -r command. If you use -g, you must include all the taxa in the tree in your constrained newick string, otherwise they're unconstrained and fall out wherever. If you use -r, use need only provide a backbone tree (not all the taxa), and the remainder are fit into the tree initially via maximum parsimony, and then optimized using maximum likelihood. Here, I'll use the command -r. 
2. Run RAxML for 100 (or however many you want) searches for the best scoring constrained tree
```{bash eval=FALSE}
[RAxML] -T 8 -f d -N 100 -m [...] -q [...] -s [...] -r [constrained newick tree] -n [...]
```
3. Run RAxML for 1000 (or however many you want) bootstraps
```{bash eval=FALSE}
[RAxML] -T 8 -f d -# 1000 -m [...] -q [...] -s [...] -n [...] -b 12345
```
4. Add your 1000 bootstraps to your best constrained tree
```{bash eval=FALSE}
[RAxML] -T 8 -f b -m [...] -q [...] -s [...] -z [boostraps, step2] -t [best tree, step1] -n [...]
# -f h will give you the likelihood of each bootstrap given your provided (constrained) tree
```

## Compiling the Trees and Comparing Them
1. When you're done with all the constrained trees, open the files and paste each tree into a new text file which contains all of the topologies (constrained and unconstrained)  
    +keep track of which tree is in which position (1 tests X,Y; 2 tests A,B; etc)
```{r eval=FALSE}
((X,Y),A,B,C,D)
((A,B),X,Y,C,D)
(((A,B),(X,Y)),C,D)
```
2. Run RAxML to get the log likelihood values between all sets of trees (command -f h), this is for your own use, or inclusion in a table
```{bash eval=FALSE}
[RAxML] -T 8 -f d -N 100 -m [...] -q [...] -s [...] -z [new tree, step1] -t [unconstrained tree] -n [...]
```
3. Run RAxML for the log likelihood output files in PUZZLE format, to be tested in CONSEL (command -f g)
```{bash eval=FALSE}
[RAxML] -T 8 -f g -m GTRGAMMA -q [...] -s [...] -z [new tree, step1] -n [...]
```
4. Compile the CONSEL files:
```{bash eval=FALSE}
#navigate to your CONSEL directory
cd src
make 
make install
```
5. Navigate back to the folder encapsulating CONSEL (cd ..) and enter your newly named log likelihood file from step3
```{bash eval=FALSE}
./consel/bin/makermt --puzzle [drag/drop the file from step3]
```
6. You've now got a [name.rmt] file which can be used by CONSEL, so run it:
```{bash eval=FALSE}
./consel/bin/consel [drag/drop the file from step5]
```
7. You should now have two files, [name.pv] and [name.ci]. The .pv file is all we care about, so read your results in the program catpv:
```{bash eval=FALSE}
./consel/bin/catpv [drag/drop the .pv file from step6]
```
8. BOOM. You should be sorted.
